---
title: "Paper 1 -- Data Management (R pipeline)"
subtitle: "Are informality traps real? Ghana AHIES 2022-2024"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  word_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
PROJECT_ROOT <- here::here()
knitr::opts_knit$set(root.dir = PROJECT_ROOT)
setwd(PROJECT_ROOT)
library(dplyr)
library(ggplot2)
library(flextable)
source("R/pipeline.R")
```

## Purpose

This document builds the analysis-ready panels from the three raw AHIES
annual extracts (2022, 2023, 2024 -- three **independent** within-year
panels, not a single long panel; no individual is linked across years),
and reports the five feasibility gates that decide whether the analysis
is informal- or vulnerable-employment-led.

Every constructed column in this R pipeline was cross-validated
column-by-column against an independent Python implementation of the
same construction logic, for all three years, before being trusted for
estimation (including one real discrepancy caught and fixed: pandas'
`.transform("first")` silently skips missing values when building the
Wooldridge initial condition, which the naive R equivalent does not
replicate without an explicit fix -- see `R/dynamics.R::first_non_na()`).

## Building the panels

```{r build, cache=TRUE}
manifest <- build_and_save_all_years()
```

```{r build-summary}
summary_tbl <- data.frame(
  Year = c(2022, 2023, 2024),
  `Person-quarter rows` = sapply(c("2022","2023","2024"), function(y) manifest[[y]]$n_rows),
  `Unique persons` = sapply(c("2022","2023","2024"), function(y) manifest[[y]]$n_persons),
  check.names = FALSE
)
flextable(summary_tbl) |> autofit()
```

Pooled file: `r manifest$pooled$n_rows` person-quarter rows across all three years (`data/processed_r/pooled_person_quarter.csv`).

## Feasibility gates

```{r gates-table}
gate_rows <- list()
for (year in c(2022, 2023, 2024)) {
  g <- manifest[[as.character(year)]]$gates
  gate_rows[[length(gate_rows) + 1]] <- data.frame(
    Year = year,
    Gate = c("1. Labour universe", "2. >=3 quarters", "3. Informality route", "4. Panel linkage", "5. Earnings quality"),
    Evidence = c(
      sprintf("%.1f%% missing econact among working-age", 100 * g$gate1_labour_universe$overall_missing_rate),
      sprintf("%d / %d persons (%.1f%%)", g$gate2_within_year_balance$n_persons_ge3_quarters,
              g$gate2_within_year_balance$n_unique_persons, 100 * g$gate2_within_year_balance$share_ge3_quarters),
      sprintf("has_contract missing %.1f%% within employees", 100 * g$gate3_informality_route$contract_missing_rate_within_employees),
      sprintf("%.1f%% sex-constant across linked quarters", 100 * g$gate4_panel_linkage$share_multiquarter_sex_constant),
      sprintf("%.1f%% missing among all employed; %.1f%% among employees only",
              100 * g$gate5_earnings_quality$missing_rate_among_all_employed,
              100 * g$gate5_earnings_quality$missing_rate_among_employees_only)
    )
  )
}
gates_tbl <- bind_rows(gate_rows)
flextable(gates_tbl) |> autofit() |> merge_v(j = "Year")
```

**Decision (Gate 3):** `informal_it` is built from `has_contract` +
`soc_sec_entitled` for employees, and `inst_sector` (Private Informal) +
family-worker status for everyone else employed. `ssnit_flag` is excluded
from the headline rule (too sparse -- see the gate-3 evidence above) and
kept only as a robustness-only cut.

## Dependent-variable base rates

```{r dv-rates, fig.width=7, fig.height=4}
rates <- list()
for (year in c(2022, 2023, 2024)) {
  df <- readr::read_csv(sprintf("data/processed_r/%d_person_quarter_unbalanced.csv", year), show_col_types = FALSE)
  emp <- df %>% filter(employed_it == TRUE)
  rates[[length(rates) + 1]] <- data.frame(
    year = year,
    dv = c("informal_it", "vulnerable_it"),
    rate = c(mean(emp$informal_it, na.rm = TRUE), mean(emp$vulnerable_it, na.rm = TRUE))
  )
}
rates_df <- bind_rows(rates)

ggplot(rates_df, aes(x = factor(year), y = rate, fill = dv)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = scales::percent(rate, accuracy = 0.1)), position = position_dodge(width = 0.9), vjust = -0.4, size = 3) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_fill_manual(values = c("informal_it" = "#4C72B0", "vulnerable_it" = "#DD8452"),
                     labels = c("Informal (ILO job-based)", "Vulnerable (ICSE own-account + family)")) +
  labs(title = "Dependent-variable base rates among the employed", x = "Year", y = "Share of employed", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

`informal_it`'s ~93-95% base rate is extremely skewed; `vulnerable_it`'s
~74-76% is far better balanced for identifying quarter-to-quarter movement
-- the basis for the Step-0 DV-lead decision in the results report.

## Known open items

Known scope decisions, none of which change any reported estimate:
no explicit design-stratum variable in the raw extract (`region x urbrur`
used as a proxy); SSNIT-specific informality route available only as a
robustness cut; secondary-job informality not constructed (main job only,
standard practice in this literature).
