---
title: "Are Informality Traps Real?"
subtitle: "State dependence in informal and vulnerable employment across Ghana's cost-of-living crisis, 2022-2024 -- full R estimation record"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    df_print: default
    theme: flatly
    fig_caption: true
    mathjax: null
---

```{r setup, include=FALSE}
# Full transparent record: code visible (foldable, shown by default)
# throughout, every table/figure/diagnostic rendered in the same order as
# the manuscript, nothing summarized away. Self-contained HTML (figures
# base64-embedded by rmarkdown/pandoc automatically under
# self_contained: true) is the sole output format for this report.
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 7, fig.height = 4.2)
PROJECT_ROOT <- here::here()
library(dplyr)
library(tidyr)
library(ggplot2)
library(flextable)
library(jsonlite)
library(DT)

show_table <- function(df, caption = NULL, digits = 4, page_length = 15) {
  DT::datatable(df, caption = caption, rownames = FALSE, filter = "top",
                 extensions = "Buttons",
                 options = list(pageLength = page_length, scrollX = TRUE,
                                dom = "Bfrtip", buttons = c("copy", "csv"))) |>
    DT::formatRound(columns = which(sapply(df, is.numeric)), digits = digits)
}

# absolute path -- knitr resets the working directory to the Rmd's own
# folder between every chunk (see the "in_dir" warning), so a setwd() in
# this chunk alone does not persist for later chunks. PROJECT_ROOT itself
# is resolved once above via here::here(), so this whole document runs
# unchanged from a fresh clone on any machine.
TAB <- file.path(PROJECT_ROOT, "output", "tables_r")
pal_dv <- c(informal_it = "#4C72B0", vulnerable_it = "#DD8452")
dv_label <- c(informal_it = "Informal (ILO job-based)", vulnerable_it = "Vulnerable (ICSE)")
theme_set(theme_minimal(base_size = 11))

rj <- function(path) jsonlite::fromJSON(path, simplifyVector = FALSE)
num <- function(x) as.numeric(x)
```

This report follows the same order as the manuscript's Data and Methods
and Results and Discussion sections (§3-4): setup and sample construction
first, then descriptives, the core state-dependence estimate, the crisis
comparison, heterogeneity, the inflation/entry analysis, and finally the
diagnostics and robustness battery. It carries substantially more detail
than the manuscript at every step -- full parameter vectors (not just the
headline lag), every fit statistic, formal interaction tests behind every
subgroup figure, and the exact numeric decompositions behind each
reconciliation claim -- with the code that produced each number shown
alongside it (see the "Code" buttons throughout).

## 0. Setup, sample construction, and feasibility gates

### 0.1 Methodology note

All estimation in this report was run in R (this machine has no Stata). The
core specification -- a dynamic random-intercept probit with the
Wooldridge (2005) initial-conditions device -- is fit by a custom
Gauss-Hermite quadrature maximum-likelihood engine (`R/re_probit.R`):
analytic score, `ucminf` optimizer, cluster-robust (EA) sandwich standard
errors built from the same per-person score contributions. This engine was
cross-validated to 6 significant figures against an independently-built
Python implementation of the identical model before being trusted.
`lme4::glmer` was tried first and did not return after 47+ minutes of
CPU time on this exact specification (39 parameters, ~45,000 observations,
one scalar random intercept) -- apparently a pathological interaction with
this data's many region/sector dummy columns, not merely slowness -- so it
was abandoned in favour of the custom engine. `brms` (Stan backend) was
confirmed to work on this machine and was run for a single flagship
cross-check model (§6.2) per the project's own recommendation to run the
core estimate in one engine and cross-check beta in another -- but a fully
converged fit on this specification (39 fixed effects, ~45,000
observations, a hierarchical structure with ~19,500 person-level
intercepts) proved too expensive for a single R session (a 2-chain,
1200-iteration run was on track for roughly two hours before being stopped
in favour of a bounded, single-chain, capped-tree-depth run). §6.2 reports
that bounded run's result honestly, including that it did not fully
converge, rather than presenting it as a clean second estimate.

Average partial effects (APEs) use the closed-form population-averaged
probability implied by a normal random intercept -- integrating a normal
random intercept alpha (variance sigma_alpha^2) out of Phi(x'beta + alpha)
gives Phi( x'beta / sqrt(1 + sigma_alpha^2) ) -- evaluated as a discrete
difference (regressor swapped between two values for every observation,
held at its own other covariate values, averaged).
For any binary (0/1) regressor in a plain probit comparison model, this
discrete-difference definition is used throughout -- **not** the continuous
partial derivative that `marginaleffects::avg_slopes()` (and Python's
`statsmodels.get_margeff()`) compute by default for a variable not
explicitly flagged as a dummy. This distinction was caught during
development: it changes a reported average marginal effect by roughly 50%
on this data (0.40 discrete vs 0.26 continuous-derivative, same fitted
model) and is not a rounding difference -- it is the difference between the
right quantity and the wrong one for a 0/1 regressor.

The simplified Heckman (1981) reduced-form approximation that appeared in
an earlier draft of this analysis is **dropped** from this report: it
degenerated (the random-intercept variance collapsed to the optimizer
boundary, σ_α ≈ 1e-7) and produced an implausible, unusable estimate. The
Mundlak/CRE pooled-probit cross-check (§6.4) is retained as the estimator
comparison.

### 0.2 Data load

```{r load-core}
core_summary <- readr::read_csv(file.path(TAB, "core_models_summary.csv"), show_col_types = FALSE)
step0 <- readr::read_csv(file.path(TAB, "step0_dv_lead_decision.csv"), show_col_types = FALSE)
```

### 0.3 Feasibility gates

Five gates (per year) decide whether the panel supports the design at
all and which DV leads: a per-quarter labour universe check, within-year
panel-balance, the informality-route decision, panel-linkage validity,
and earnings-quality. This bundle's data-inspection stage that originally
produced a written record of that evidence (`docs/gates_report.md`) isn't
included here -- if you produce an equivalent file at that path (one
`## <year>` heading per year, each followed by a fenced ```json block
with `gate1_labour_universe`, `gate2_within_year_balance`,
`gate3_informality_route`, `gate4_panel_linkage`, `gate5_earnings_quality`
keys), this section will parse and render it; otherwise it renders a
short note and every other section is unaffected.

```{r feasibility-gates-load}
gates_path <- file.path(PROJECT_ROOT, "docs", "gates_report.md")
gates <- list()
if (file.exists(gates_path)) {
  gates_lines <- readLines(gates_path)
  year_blocks <- list()
  current_year <- NA_integer_
  buf <- character(0)
  in_json <- FALSE
  for (ln in gates_lines) {
    if (grepl("^## [0-9]{4}$", ln)) {
      current_year <- as.integer(trimws(sub("^## ", "", ln)))
    } else if (grepl("^```json", ln)) {
      in_json <- TRUE; buf <- character(0)
    } else if (grepl("^```\\s*$", ln) && in_json) {
      in_json <- FALSE
      year_blocks[[as.character(current_year)]] <- paste(buf, collapse = "\n")
    } else if (in_json) {
      buf <- c(buf, ln)
    }
  }
  gates <- lapply(year_blocks, function(x) jsonlite::fromJSON(x, simplifyVector = FALSE))
}
```

```{r feasibility-gates-table}
if (length(gates) > 0) {
  gate_rows <- lapply(names(gates), function(yr) {
    g <- gates[[yr]]
    data.frame(
      Year = as.integer(yr),
      `Gate 1 (labour universe)` = ifelse(g$gate1_labour_universe$pass, "PASS", "FAIL"),
      `N working-age person-qtrs` = g$gate1_labour_universe$n_working_age_rows,
      `Gate 2 (within-year balance)` = ifelse(g$gate2_within_year_balance$pass, "PASS", "FAIL"),
      `Share >=3 quarters` = scales::percent(g$gate2_within_year_balance$share_ge3_quarters, accuracy = 0.1),
      `Share all 4 quarters` = scales::percent(g$gate2_within_year_balance$share_all4_quarters, accuracy = 0.1),
      `Gate 4: N persons` = g$gate4_panel_linkage$n_unique_persons,
      check.names = FALSE
    )
  })
  flextable(bind_rows(gate_rows)) |> autofit()
} else {
  cat("docs/gates_report.md not found -- see the note above for the expected format; re-knit this report to see the gate table once it exists.")
}
```

**Gate 3 (informality-route) and Gate 5 (earnings quality)** are decisions,
not pass/fail booleans -- quoted directly from the recorded evidence
(identical across years, so shown once):

```{r feasibility-gates-narrative, results='asis'}
if (length(gates) > 0) {
  g1 <- gates[[1]]
  cat(sprintf("- **Gate 3 decision:** %s\n\n", g1$gate3_informality_route$decision))
  cat(sprintf("- **Gate 5 verdict:** %s\n\n", g1$gate5_earnings_quality$verdict))
} else {
  cat("Pending -- see the note above.\n\n")
}
```

All five gates pass in every year; none forced a design change beyond the
one already anticipated in the plan (§9.3: switch lead DV to
`vulnerable_it` if the informality route proves thin -- confirmed
empirically in §1 below, not because Gate 3 itself failed).

### 0.4 Sample construction summary

The funnel from the raw per-quarter labour universe (Gate 1) down to the
estimation universe each core model actually fits on (employed, lag
defined -- i.e. not a person's first observed quarter):

```{r sample-construction}
core_n_display <- core_summary %>%
  transmute(Year = year, DV = dv_label[dv],
            `Estimation N (lag defined)` = format(n_obs, big.mark = ","),
            `At-risk N (lag=0)` = format(n_lag0, big.mark = ","))

if (length(gates) > 0) {
  sample_funnel <- bind_rows(lapply(names(gates), function(yr) {
    g <- gates[[yr]]
    data.frame(year = as.integer(yr),
               n_working_age_pq = g$gate1_labour_universe$n_working_age_rows,
               n_employed_pq = g$gate3_informality_route$n_employed)
  })) %>%
    left_join(core_summary %>% filter(dv == "vulnerable_it") %>%
                transmute(year, n_obs_vulnerable = n_obs, n_lag0_vulnerable = n_lag0), by = "year") %>%
    left_join(core_summary %>% filter(dv == "informal_it") %>%
                transmute(year, n_obs_informal = n_obs, n_lag0_informal = n_lag0), by = "year")

  funnel_display <- sample_funnel %>%
    transmute(Year = year,
              `Working-age person-qtrs (Gate 1)` = format(n_working_age_pq, big.mark = ","),
              `Employed person-qtrs (Gate 3)` = format(n_employed_pq, big.mark = ","),
              `Estimation N, vulnerable_it (lag defined)` = format(n_obs_vulnerable, big.mark = ","),
              `At-risk N, vulnerable_it (lag=0)` = format(n_lag0_vulnerable, big.mark = ","),
              `Estimation N, informal_it (lag defined)` = format(n_obs_informal, big.mark = ","),
              `At-risk N, informal_it (lag=0)` = format(n_lag0_informal, big.mark = ","))
  flextable(funnel_display) |> autofit()
} else {
  cat("The raw-universe and employed-universe ends of the funnel come from docs/gates_report.md, which isn't present (see the note above) -- showing only the estimation-universe end, from core_models_summary.csv:\n\n")
  flextable(core_n_display) |> autofit()
}
```

The estimation universe (~44,700-49,600 person-quarters/year) is smaller
than the raw working-age universe mainly because the dynamic model needs
the lag -- a person's first observed quarter in a year contributes to the
Wooldridge initial condition but not to the likelihood's t=2,3,4 terms.
Construction rules for every column (the ILO job-based `informal_it`
rule, the ICSE `vulnerable_it` rule, lags, and the Wooldridge initial
condition) are in `R/construct.R` and `R/dynamics.R`, commented inline.

## 1. Descriptive patterns and the choice of outcome

```{r step0-table}
step0_display <- step0 %>%
  transmute(Year = year, DV = dv_label[dv],
            `Base rate` = scales::percent(base_rate_among_employed, accuracy = 0.1),
            `P(y=1|lag=1)` = round(P_y1_given_lag1, 3),
            `P(y=1|lag=0)` = round(P_y1_given_lag0, 3),
            `Raw gap` = round(raw_gap, 3),
            `N at lag=0` = n_lag0)
flextable(step0_display) |> autofit()
```

```{r step0-chart}
ggplot(step0, aes(x = factor(year), y = base_rate_among_employed, fill = dv)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = scales::percent(base_rate_among_employed, accuracy = 0.1)),
            position = position_dodge(width = 0.9), vjust = -0.4, size = 3) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_fill_manual(values = pal_dv, labels = dv_label) +
  labs(title = "DV base rates among the employed", x = "Year", y = NULL, fill = NULL) +
  theme(legend.position = "bottom")
```

**Decision: `vulnerable_it` leads.** Its ~74-76% base rate is far less
skewed than `informal_it`'s ~93-95%, giving 3x the "at-risk" (lag=0)
person-quarters to identify entry (e.g. 2022: 12,029 vs 3,763) -- the
sharper-identification criterion the analysis plan set out in advance.
`informal_it` is reported throughout as the co-primary DV, since it is the
paper's own named object.

### 1.1 Four-state transition matrices

Quarter-to-quarter transitions among the four labour-force states
(`formal_employed`, `informal_vulnerable_employed`, `unemployed`,
`out_of_lf`; plan §7.1), **adjacent quarters only** -- a skipped quarter is
never counted as a transition (`R/transitions.R::quarter_pairs()` enforces
`prev_t == t - 1`). Each cell shows the row-normalized transition
probability with the underlying (person-weighted) raw count in
parentheses; rows are the state at *t-1*, columns the state at *t*.

```{r transitions-load}
STATE_ORDER <- c("formal_employed", "informal_vulnerable_employed", "unemployed", "out_of_lf")
state_label <- c(formal_employed = "Formal employed",
                  informal_vulnerable_employed = "Informal/vulnerable employed",
                  unemployed = "Unemployed", out_of_lf = "Out of LF")

load_transition_year <- function(y) {
  tm <- readr::read_csv(file.path(TAB, sprintf("%d_transition_matrix.csv", y)), show_col_types = FALSE)
  gf <- readr::read_csv(file.path(TAB, sprintf("%d_gross_flows.csv", y)), show_col_types = FALSE)
  tm <- tm %>% arrange(match(prev_status, STATE_ORDER))
  gf <- gf %>% arrange(match(prev_status, STATE_ORDER))
  combo <- data.frame(prev_status = tm$prev_status)
  for (s in STATE_ORDER) {
    combo[[s]] <- sprintf("%s%%\n(n=%s)",
                           formatC(100 * tm[[s]], format = "f", digits = 1),
                           format(gf[[s]], big.mark = ","))
  }
  combo$prev_status <- state_label[combo$prev_status]
  names(combo)[-1] <- state_label[STATE_ORDER]
  combo
}
```

```{r transitions-2022}
flextable(load_transition_year(2022)) |> autofit() |>
  set_caption("2022 transition matrix (row = state at t-1, column = state at t)")
```

```{r transitions-2023}
flextable(load_transition_year(2023)) |> autofit() |>
  set_caption("2023 transition matrix")
```

```{r transitions-2024}
flextable(load_transition_year(2024)) |> autofit() |>
  set_caption("2024 transition matrix")
```

**Reading the diagonal and off-diagonal pattern:** the diagonal is the
dominant entry in every row and every year. `informal_vulnerable_employed`
is the stickiest state throughout (78.4% quarter-to-quarter persistence in
2022, rising to 88.5% in 2023 and 88.0% in 2024). The off-diagonal cells
are informative about direction, not just magnitude: in every year, the
*probability* of moving from unemployment into informal/vulnerable work
(32.5% in 2022, 22.2% in 2023, 24.3% in 2024) is far larger than the
reverse -- leaving informal work for unemployment (6.1%, 3.8%, 2.9%) -- and
the same asymmetry, smaller but consistent, holds between
`out_of_lf` and informal work (22.2% in vs. 13.2% out in 2022; the gap
narrows but does not reverse by 2024). Informal/vulnerable employment is
therefore a much easier state to fall into than to leave, in every year of
the panel -- the raw, uncontrolled version of the "trap" pattern the
dynamic probit in §2 tests once heterogeneity is removed.

Every state's persistence rises sharply from 2022 to 2023-2024, not just
`informal_vulnerable_employed`'s: formal employment's diagonal goes
63.4% -> 86.7% -> 63.4%, unemployment's goes 22.9% -> 56.5% -> 60.4%, and
`out_of_lf`'s goes 68.5% -> 80.5% -> 83.3%. 2022 was a high-churn year
across *every* labour-force state, not selectively for informal work --
consistent with the crisis-comparison finding in §3 that state dependence
is lowest in 2022 and highest in 2023, and a first, purely descriptive
hint that the crisis year itself was too chaotic for any state (including
the "trap") to fully consolidate.

### 1.2 Full covariate descriptives by year and vulnerable-employment status

Means (numeric covariates) and shares (categorical covariates), by year
and by current `vulnerable_it` status, on the same estimation universe the
core models use.

```{r descriptives-n}
desc_n <- readr::read_csv(file.path(TAB, "descriptives_n_summary.csv"), show_col_types = FALSE)
desc_n_display <- desc_n %>%
  transmute(Year = year, Status = vulnerable_it, `N person-quarters` = n_person_quarters, `N persons` = n_persons)
flextable(desc_n_display) |> autofit() |> merge_v(j = "Year")
```

```{r descriptives-table}
desc <- readr::read_csv(file.path(TAB, "descriptives_by_year_vulnerable.csv"), show_col_types = FALSE)
desc_display <- desc %>%
  mutate(value = ifelse(type == "numeric", sprintf("mean=%.3f, sd=%.3f", mean, sd), sprintf("share=%.3f", mean))) %>%
  select(year, variable, level, vulnerable_it, value, n_obs) %>%
  arrange(variable, level, year, vulnerable_it)
show_table(desc_display, caption = "Covariate descriptives by year x vulnerable_it status", page_length = 20)
```

## 2. Core state-dependence estimates: true vs. spurious persistence

```{r core-table}
core_display <- core_summary %>%
  transmute(Year = year, DV = dv_label[dv], N = n_obs, `At-risk N` = n_lag0,
            `APE(lag)` = sprintf("%.4f (%.4f)", ape_lag, ape_lag_se),
            `Naive coef` = round(naive_coef_lag, 3),
            `sigma_alpha` = sprintf("%.3f (%.3f)", sigma_alpha, se_sigma_alpha),
            `Wooldridge-aux Wald p` = ifelse(wald_aux_p < 0.0001, "<0.0001", round(wald_aux_p, 4)),
            `Unstable?` = ifelse(unstable, "YES", ""))
ft_core <- flextable(core_display) |> autofit()
if ("Unstable?" %in% names(core_display)) {
  unstable_rows <- which(core_display$`Unstable?` == "YES")
  if (length(unstable_rows)) ft_core <- ft_core |> bg(i = unstable_rows, bg = "#FDECEA")
}
ft_core
```

```{r core-chart, fig.height=4.5}
ggplot(core_summary, aes(x = factor(year), y = ape_lag, fill = dv)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = ape_lag - 1.96 * ape_lag_se, ymax = ape_lag + 1.96 * ape_lag_se),
                position = position_dodge(width = 0.8), width = 0.15) +
  scale_fill_manual(values = pal_dv, labels = dv_label) +
  labs(title = "Corrected average partial effect of state dependence, by year",
       subtitle = "Dynamic RE-probit + Wooldridge device; error bars = 95% CI",
       x = "Year", y = "APE(lag)", fill = NULL) +
  theme(legend.position = "bottom")
```

**Reading this (the core identification check):** the Wooldridge auxiliaries
are jointly significant at p<0.0001 in every model -- the device is doing
real work. The lag coefficient collapses sharply once heterogeneity is
controlled (e.g. 2022 `vulnerable_it`: naive coefficient 1.45 -> corrected
APE 0.095), the visible signature of spurious persistence being removed.
σ_α is well away from zero throughout. **`vulnerable_it` carries the
clearer, better-identified trap result** (APE 0.095-0.163 in 2022-2023,
tightly estimated); `informal_it`'s picture is noisier across years,
plausibly because its ~93-95% base rate leaves few "at risk" observations
per year to pin the estimate down (§6.1 below flags this formally).

**Small-T caveat (Akay 2012).** Every model in this report is estimated on
a T=4-quarter panel (T=2-4 effective, since the estimation universe needs
the lag). The Wooldridge (2005) device used throughout is the standard
fix for the initial-conditions problem in short panels, but it is a
large-N/fixed-T asymptotic device, not a small-sample-exact one --
Akay (2012) shows via Monte Carlo that dynamic RE-probit bias under this
device is not negligible at T as short as this project's, particularly for
σ_α and, to a lesser extent, β. Nothing in the diagnostics below
contradicts the headline results, but the AME magnitudes -- especially for
the DV/year cells flagged unstable in §6.1 -- should be read as
identified up to this small-T bias, not as exact.

### 2.1 Full coefficient tables and fit statistics

Every fitted parameter (not just the lag) for all six core models: the
lag β, the Wooldridge initial condition (`dv_i1`), every covariate γ, the
within-person means and sector shares (the Wooldridge/Mundlak auxiliary
terms), the quarter dummies, and σ_α, all with cluster-robust (EA)
sandwich SEs. Main text: `vulnerable_it` (the lead DV). Companion:
`informal_it`. Both tables carry a footer row with the **corrected
discrete lag APE** (§0.1's closed-form Φ(x'β/√(1+σ_α²)) contrast) -- this
is *not* the same quantity as the raw `L_dv` coefficient a few rows above
it, and the two should never be read interchangeably (the plain
coefficient is on the probit index scale; the APE is a probability-scale
population-averaged effect).

```{r fit-stats-table}
fit_stats <- readr::read_csv(file.path(TAB, "fit_statistics.csv"), show_col_types = FALSE)
fit_stats_display <- fit_stats %>%
  transmute(Year = year, DV = dv_label[dv], `k (params)` = k_params,
            `Log-lik` = round(loglik, 2), AIC = round(aic, 2),
            `BIC (N=person-quarters)` = round(bic_n_obs, 2),
            `BIC (N=persons)` = round(bic_n_persons, 2),
            `N (person-qtrs)` = n_obs, `N (persons)` = n_persons,
            `sigma_alpha` = sprintf("%.4f (%.4f)", sigma_alpha, se_sigma_alpha),
            Converged = converged)
flextable(fit_stats_display) |> autofit()
```

BIC is reported on both candidate sample sizes (person-quarters and
persons) rather than picking one silently -- for a random-intercept model
the "N" BIC should use is genuinely ambiguous in the literature; both are
given so either convention can be applied downstream.

```{r full-coef-vulnerable}
fc_vul <- readr::read_csv(file.path(TAB, "full_coef_vulnerable_it_long.csv"), show_col_types = FALSE)
footer_vul <- core_summary %>% filter(dv == "vulnerable_it") %>%
  transmute(year, param = "APE(lag) [corrected, discrete, Phi(x'b/sqrt(1+sa^2))]",
            coef = ape_lag, se = ape_lag_se, z = NA_real_, p = NA_real_)
fc_vul_display <- bind_rows(fc_vul, footer_vul) %>% arrange(year, param != "APE(lag) [corrected, discrete, Phi(x'b/sqrt(1+sa^2))]")
show_table(fc_vul_display, caption = "vulnerable_it: full coefficient vector, all years (footer row = corrected APE, not a raw coefficient)", page_length = 43)
```

```{r full-coef-informal}
fc_inf <- readr::read_csv(file.path(TAB, "full_coef_informal_it_long.csv"), show_col_types = FALSE)
footer_inf <- core_summary %>% filter(dv == "informal_it") %>%
  transmute(year, param = "APE(lag) [corrected, discrete, Phi(x'b/sqrt(1+sa^2))]",
            coef = ape_lag, se = ape_lag_se, z = NA_real_, p = NA_real_)
fc_inf_display <- bind_rows(fc_inf, footer_inf) %>% arrange(year, param != "APE(lag) [corrected, discrete, Phi(x'b/sqrt(1+sa^2))]")
show_table(fc_inf_display, caption = "informal_it: full coefficient vector, all years (companion; footer row = corrected APE)", page_length = 43)
```

## 3. Crisis comparison: is state dependence higher in the crisis year?

```{r rq3-load}
rq3_inf <- rj(file.path(TAB, "rq3_pooled_informal_it_diagnostics.json"))
rq3_vul <- rj(file.path(TAB, "rq3_pooled_vulnerable_it_diagnostics.json"))

rq3_rows <- function(x) {
  data.frame(
    dv = x$dv,
    year = c(2022, 2023, 2024),
    ape = sapply(x$ape_by_year, function(a) num(a$ape)),
    se = sapply(x$ape_by_year, function(a) num(a$se))
  )
}
rq3_df <- bind_rows(rq3_rows(rq3_inf), rq3_rows(rq3_vul))
```

```{r rq3-chart}
ggplot(rq3_df, aes(x = factor(year), y = ape, color = dv, group = dv)) +
  geom_line(linewidth = 1) + geom_point(size = 2.5) +
  geom_errorbar(aes(ymin = ape - 1.96*se, ymax = ape + 1.96*se), width = 0.1) +
  scale_color_manual(values = pal_dv, labels = dv_label) +
  labs(title = "Pooled state-dependence APE by year (year x lag interaction)",
       subtitle = "Person/EA keys re-qualified by year before pooling",
       x = "Year", y = "APE(lag)", color = NULL) +
  theme(legend.position = "bottom")
```

```{r rq3-wald}
wald_tbl <- data.frame(
  DV = c(dv_label["informal_it"], dv_label["vulnerable_it"]),
  `chi2` = c(round(num(rq3_inf$wald_H0_beta_equal_across_years$stat), 2),
             round(num(rq3_vul$wald_H0_beta_equal_across_years$stat), 2)),
  df = c(rq3_inf$wald_H0_beta_equal_across_years$df, rq3_vul$wald_H0_beta_equal_across_years$df),
  p = c("<0.0001", "<0.0001"), check.names = FALSE
)
flextable(wald_tbl) |> autofit()
```

**The exact contrast being tested.** The pooled specification above is
`y_it ~ ... + L_dv + year_2023 + year_2024 + year_2023:L_dv + year_2024:L_dv + ...`,
with 2022 as the reference year. The lag effect in 2022 is the base
coefficient on `L_dv` alone; in 2023 it is `L_dv + year_2023:L_dv`; in 2024
it is `L_dv + year_2024:L_dv`. The test above is **H0: year_2023_x_lag = 0
AND year_2024_x_lag = 0** -- a joint 2-restriction Wald test on exactly
those two interaction coefficients
(`R/re_probit.R::wald_test(fit, c("year_2023_x_lag", "year_2024_x_lag"))`),
using the model's cluster-robust variance-covariance submatrix for those
two parameters:

```{r chisq-sanity, results='asis'}
rq3_vul_wald <- rq3_vul$wald_H0_beta_equal_across_years
cat(sprintf("Test object (`R/re_probit.R::wald_test`), `vulnerable_it`: parameters tested = c(\"year_2023_x_lag\", \"year_2024_x_lag\"). stat=%.3f, df=%d, p=%.6g\n\n",
            num(rq3_vul_wald$stat), rq3_vul_wald$df, num(rq3_vul_wald$p)))
rq3_inf_wald <- rq3_inf$wald_H0_beta_equal_across_years
cat(sprintf("Test object, `informal_it`: parameters tested = c(\"year_2023_x_lag\", \"year_2024_x_lag\"). stat=%.3f, df=%d, p=%.6g\n",
            num(rq3_inf_wald$stat), rq3_inf_wald$df, num(rq3_inf_wald$p)))
```

**df=2 is correct and corresponds to the two year-interaction terms being
restricted, not to the year main effects** (`year_2023`, `year_2024`
alone) -- those main effects capture a level shift in the *intercept* by
year (baseline prevalence), a completely different, untested hypothesis;
they do not "load through" the lag-interaction test in any way, because
they are separate columns in the design matrix with their own separate
coefficients, entered simultaneously but not restricted by this particular
Wald test. Rejecting H0 at p<0.0001 means: the *effect of the lag* on the
outcome differs across years, which is precisely "state dependence differs
by year."

**Finding: in this pooled specification, state dependence peaks in 2023,
not the crisis year 2022, for both DVs.** This does not match the "crisis
deepened the trap" hypothesis in the project overview -- it is a genuine
departure from that prior, not a confirmation, and needs new framing text
rather than a number swap. A plausible reading: 2022 was too
chaotic/high-churn for the trap to fully consolidate (§1.1's transition
matrices already show this descriptively -- *every* state's persistence
rises from 2022 to 2023-2024, not just informal/vulnerable employment's);
2023 -- the IMF-programme/adjustment year -- is where people who fell into
vulnerable employment *stuck*. H0 (β equal across years) is rejected at
p<0.0001 for both DVs.

### 3.1 Reconciling the per-year and pooled `informal_it` profile

**This "for both DVs" claim needs one explicit qualification: it is true
of this pooled, year-interaction specification, but it is *not* true of
the separate per-year core models in §2**, where `informal_it` peaks in
2024 (APE 0.174), not 2023 (APE 0.022). §6.1 already flags that 2024
`informal_it` cell as unstable (thin at-risk sample, low σ_α) -- this
subsection puts the two `informal_it` profiles side by side and gives the
exact numeric decomposition of why they diverge.

```{r rq3-reconcile}
core_informal <- core_summary %>% filter(dv == "informal_it") %>%
  transmute(year, spec = "Per-year core model (§2, separate sigma_alpha each year)",
            ape = ape_lag, se = ape_lag_se)
pooled_informal <- rq3_df %>% filter(dv == "informal_it") %>%
  transmute(year, spec = "Pooled crisis-comparison model (§3, single sigma_alpha across years)", ape, se)
reconcile_df <- bind_rows(core_informal, pooled_informal)

reconcile_display <- reconcile_df %>%
  transmute(Year = year, Specification = spec, `APE(lag)` = sprintf("%.4f (%.4f)", ape, se))
flextable(reconcile_display) |> autofit() |> merge_v(j = "Year")
```

```{r rq3-reconcile-chart, fig.height=3.8}
ggplot(reconcile_df, aes(x = factor(year), y = ape, color = spec, group = spec)) +
  geom_line(linewidth = 1) + geom_point(size = 2.5) +
  geom_errorbar(aes(ymin = ape - 1.96*se, ymax = ape + 1.96*se), width = 0.1) +
  scale_color_manual(values = c("Per-year core model (§2, separate sigma_alpha each year)" = "#4C72B0",
                                 "Pooled crisis-comparison model (§3, single sigma_alpha across years)" = "#C44E52")) +
  labs(title = "informal_it: per-year vs. pooled state-dependence profile",
       x = "Year", y = "APE(lag)", color = NULL) +
  theme(legend.position = "bottom", legend.text = element_text(size = 8))
```

The two profiles agree that 2022 is the low point and 2023 is high, but
disagree on 2024: the per-year model puts 2024 highest of all
(0.174), the pooled model puts it in the middle (0.133), below 2023
(0.215). The per-year 2024 `informal_it` estimate is built on a thin,
flagged-unstable at-risk sample (§6.1) and swings on comparatively little
identifying variation -- that part of the story stands. **It is NOT
primarily the pooled model's single σ_α "pulling every year toward a
common scale."** Formally decomposing the per-year-vs-pooled APE gap into
a σ_α-swap component and a β-swap component -- same closed-form APE,
evaluated at the per-year β with the pooled σ_α, and vice versa, on
identical data, no refitting -- shows the exact mechanism:

```{r reconcile-table}
recon <- readr::read_csv(file.path(TAB, "peryear_vs_pooled_reconciliation.csv"), show_col_types = FALSE)
recon_display <- recon %>%
  transmute(Year = year, DV = dv_label[dv],
            `sigma_a (per-year)` = round(sigma_alpha_per_year, 3),
            `sigma_a (pooled)` = round(sigma_alpha_pooled, 3),
            `Deflation (per-year)` = round(deflation_per_year, 3),
            `Deflation (pooled)` = round(deflation_pooled, 3),
            `APE per-year` = round(ape_per_year_actual, 4),
            `APE pooled slice` = round(ape_pooled_slice, 4),
            `Total gap` = round(total_gap, 4),
            `% from sigma_a swap` = sprintf("%.1f%%", pct_gap_from_sigma),
            `% from beta swap` = sprintf("%.1f%%", pct_gap_from_beta))
show_table(recon_display, caption = "Per-year vs. pooled APE gap, decomposed", page_length = 6)
```

**Diagnosis:** across all six DV-year cells, the σ_α-swap component
explains essentially none of the gap (-7% to +3%, i.e. often the wrong
sign) while the β-swap component explains essentially all of it (98% to
125%). The pooled model's single σ_α is **not** the mechanism -- pooling's
binding constraint is that it forces every covariate coefficient γ to be
identical across 2022-2024 (only the intercept and lag shift by year, via
the interaction terms), whereas the per-year models let sector, region,
education, household composition, etc. effects vary freely by year. For a
noisily-estimated per-year cell like 2024 `informal_it`, that re-identified
γ moves the fitted response surface far more than integrating over a
different σ_α does. `vulnerable_it`'s per-year and pooled σ_α (0.81-1.15
vs. 0.975) are far closer together than `informal_it`'s (0.35-1.58 vs.
0.499), consistent with `vulnerable_it` not needing this diagnosis in
practice. **Read the pooled profile as the more trustworthy summary of
where state dependence is highest across years** -- it is not immune to
the 2024 cell's thinness, but it is far less exposed to it than the
per-year estimate is, for the reason just given, not the σ_α-averaging
reason one might guess at first. `vulnerable_it`, which has no flagged
cell, tells the same story (2023 highest) in both specifications and does
not need this caveat.

## 4. Heterogeneity in the trap

```{r rq4-load}
rq4_files <- list.files(TAB, pattern = "^rq4_.*\\.json$", full.names = TRUE)
rq4_all <- bind_rows(lapply(rq4_files, function(f) {
  x <- rj(f)
  sg <- x$subgroup_ape
  data.frame(year = x$year, group = x$group,
             level = names(sg),
             ape = sapply(sg, function(a) num(a$ape)),
             se = sapply(sg, function(a) num(a$se)),
             wald_p = num(x$wald_H0_no_heterogeneity$p))
}))
```

```{r rq4-chart, fig.height=8}
group_labels <- c(sex = "Sex", age_band = "Age band", edu_cat = "Education", urbrur = "Locality")
rq4_all$group_label <- group_labels[rq4_all$group]

ggplot(rq4_all, aes(x = level, y = ape, fill = factor(year))) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = ape - 1.96*se, ymax = ape + 1.96*se), position = position_dodge(width = 0.8), width = 0.15) +
  facet_wrap(~group_label, scales = "free_x", ncol = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Subgroup APE(lag), vulnerable_it, by year", x = NULL, y = "APE(lag)", fill = "Year") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "bottom")
```

**Consistent pattern across all three years: urban state dependence is
roughly 2x rural.** Education shows a consistent secondary-education peak
(secondary > tertiary > basic > none in all three years) -- plausibly
workers with some schooling but not enough to credential into formal jobs
are the most "stuck." Sex differences are small; age is inconsistent across
years and not always jointly significant.

### 4.1 Formal interaction tests and full subgroup APE table

The subgroup lag-APEs behind the figure above, alongside the **formal Wald
test of the lag x characteristic interaction** (H0: no heterogeneity on
that dimension, that year) that the figure alone doesn't show a p-value
for.

```{r subgroup-wald-table}
sub_wald <- readr::read_csv(file.path(TAB, "subgroup_interaction_wald.csv"), show_col_types = FALSE)
sub_wald_display <- sub_wald %>%
  transmute(Year = year, Dimension = dimension, `chi2` = round(wald_chi2, 2), df = wald_df,
            p = ifelse(wald_p < 0.0001, "<0.0001", round(wald_p, 4)))
flextable(sub_wald_display) |> autofit()
```

```{r subgroup-ape-table}
sub_ape <- readr::read_csv(file.path(TAB, "subgroup_ape_by_dimension.csv"), show_col_types = FALSE)
sub_ape_display <- sub_ape %>%
  transmute(Year = year, Dimension = dimension, Level = level,
            `APE(lag)` = sprintf("%.4f (%.4f)", ape_lag, se))
show_table(sub_ape_display, caption = "Subgroup lag-APE by year and dimension (vulnerable_it)", page_length = 20)
```

The `age_band` interaction is the one dimension not always significant
(2023 p=0.20) -- consistent with the figure's pattern that age is the
least consistent heterogeneity dimension. `edu_cat` and `urbrur` are
jointly significant at p<0.0001 in every year; `sex` weakens to p=0.086 by
2024.

## 5. Inflation and within-year entry

```{r rq2-load}
rq2 <- rj(file.path(TAB, "rq2_event_study.json"))
entry_df <- bind_rows(lapply(rq2$entry_models, function(m) {
  data.frame(dv = m$dv, inflation_measure = m$inflation_measure, n_obs = m$n_obs,
             entry_rate = num(m$entry_rate), ame = num(m$ame_inflation_per_sd),
             se = num(m$ame_inflation_se), p = num(m$p_inflation))
}))
```

**Identification caveat, stated up front:** inflation only varies by
*quarter* within 2022 -- at most 3 independent macro data points (Q2-Q4)
identify the coefficient below, confounded with every other quarter-specific
shock. Treat these as suggestive, descriptive-causal at best.

```{r rq2-table}
entry_display <- entry_df %>%
  transmute(DV = dv_label[dv], `Inflation measure` = ifelse(inflation_measure == "inflation_yoy_qt", "YoY level", "Acceleration"),
            N = n_obs, `Entry rate` = scales::percent(entry_rate, accuracy = 0.1),
            `AME per 1 SD` = sprintf("%.4f (%.4f)", ame, se),
            p = ifelse(p < 0.0001, "<0.0001", round(p, 4)))
flextable(entry_display) |> autofit()
```

The sign is **negative** where significant -- higher/accelerating inflation
associated with *lower* quarterly entry, the opposite of the naive
"crisis pushes people into informal work" story. Given the 3-data-point
identification problem, this is most credibly read as "we cannot
distinguish an inflation effect from whatever else changed quarter to
quarter," not as evidence inflation protects against informalisation.

```{r rq2-persistence}
pers_df <- bind_rows(lapply(rq2$persistence_of_early_entrants, function(p) {
  data.frame(DV = dv_label[p$dv], `N early entrants` = p$n_early_entrants_with_later_obs,
             `Stay rate` = scales::percent(num(p$stay_rate_by_q4_or_last_obs), accuracy = 0.1), check.names = FALSE)
}))
flextable(pers_df) |> autofit()
```

Of Q2/Q3 entrants with a later 2022 observation, the majority stayed in
their new state through their last observed quarter -- a within-2022 echo
of `vulnerable_it`'s more durable persistence.

## 6. Diagnostics and robustness

### 6.1 Quadrature stability and thin-cell / separation flags

**Node-count (Gauss-Hermite quadrature) stability.** The flagship model
(2022, `vulnerable_it`) was refit at 12 (baseline), 24, 48, and 96
integration nodes; `vulnerable_it` 2023 and 2024 were each refit at 12 and
24 as a lighter confirmatory check (a full 96-node sweep on all three years
was not run -- see `R/run_node_stability.R` for the documented scope
decision).

```{r node-stability-table}
node_path <- file.path(TAB, "node_stability.csv")
if (file.exists(node_path)) {
  ns <- readr::read_csv(node_path, show_col_types = FALSE)
  ns_display <- ns %>%
    transmute(Year = year, DV = dv_label[dv], `Nodes (K)` = n_quad,
              `APE(lag)` = round(ape, 5), `sigma_alpha` = round(sigma_alpha, 4),
              `Rel. change vs K=12` = sprintf("%+.3f%%", rel_change_pct))
  ft_node <- flextable(ns_display) |> autofit() |> merge_v(j = c("Year", "DV"))
  max_rel <- max(abs(ns$rel_change_pct))
  stability_verdict <- if (max_rel < 1) {
    sprintf("The largest relative change in APE across the whole sweep is %.3f%%, well inside the plan's ~1%% target -- quadrature is not the limiting source of precision anywhere in this report.", max_rel)
  } else {
    sprintf("**Flag:** the largest relative change in APE across the sweep is %.3f%%, ABOVE the plan's ~1%% target -- treat the affected cell's APE as quadrature-sensitive, not fully settled at K=12.", max_rel)
  }
} else {
  ft_node <- NULL
  stability_verdict <- "Node-stability sweep pending."
}
ft_node
```

`r stability_verdict`

**Thin-cell / quasi-separation flags.** Every core model in §2's table
carries three stored diagnostics (`R/run_core_models.R::instability_flags()`):
at-risk (lag=0) N below 2,000; σ_α below 0.5 (well under the 0.8-1.6 range
every other core model's random-intercept SD falls in); and quasi-separation
(the population-averaged fitted probability landing within 1e-6 of 0 or 1
for any observation). Any one trips the row's "Unstable?" flag.

```{r instability-detail}
inst_display <- core_summary %>%
  filter(unstable) %>%
  transmute(Year = year, DV = dv_label[dv], `At-risk N` = n_lag0,
            `Thin at-risk (<2000)` = ifelse(thin_at_risk, "YES", "no"),
            `sigma_alpha low (<0.5)` = ifelse(sigma_alpha_near_boundary, "YES", "no"),
            `Quasi-separation` = ifelse(quasi_separation, "YES", "no"))
if (nrow(inst_display) > 0) {
  flextable(inst_display) |> autofit()
} else {
  cat("No core model tripped the instability flag.")
}
```

**The 2024 `informal_it` cell is flagged and should not be read at face
value.** Its at-risk N (see §2's table) is barely half the next-smallest
core model's, and its σ_α (0.346) is the lowest of all six core models --
both independent symptoms of a thin, sparsely-identified estimation
problem, not a genuine finding that state dependence is stronger in 2024
than in 2022-2023 for `informal_it`. **This estimate is retained in the
table above, not dropped** (per this report's own convention of flagging
rather than hiding problems), but it should not be cited as evidence about
2024 informal-employment dynamics. For 2024, and for any year-over-year
reading of `informal_it` generally, **lead with `vulnerable_it`**, which
has no flagged cell in any year and was already the better-identified DV
by the Step-0 decision in §1.

### 6.2 Bayesian (brms) cross-check

```{r brms-crosscheck, results='asis'}
brms_path <- file.path(TAB, "brms_crosscheck_2022_vulnerable_it.json")
if (file.exists(brms_path)) {
  bc <- rj(brms_path)
  cat(sprintf("Flagship model (2022, vulnerable_it) re-estimated in brms (Bayesian, Stan/NUTS): posterior mean APE(lag) = %.4f, 95%% credible interval [%.4f, %.4f], vs the quadrature-MLE point estimate %.4f (95%% CI [%.4f, %.4f]). Max Rhat = %.3f.\n\n",
              num(bc$brms_ape_mean), num(bc$brms_ape_lo), num(bc$brms_ape_hi),
              num(bc$mle_ape), num(bc$mle_ci_lo), num(bc$mle_ci_hi), num(bc$rhat_max)))

  def_match <- isTRUE(bc$definition_matches_mle_engine)
  cat(sprintf("**APE-definition check:** %s %s\n\n",
              ifelse(def_match,
                     "the brms APE above was already computed under the identical discrete closed-form contrast the MLE engine uses (Phi(x'beta/sqrt(1+sigma_alpha^2)), P(1)-P(0), per posterior draw then averaged) -- it was never routed through `marginaleffects::avg_comparisons()`/`avg_slopes()` (that path crashed on a missing package dependency and was abandoned in favour of the direct closed-form computation in `R/run_brms_ape_check.R`).",
                     "the brms APE was recomputed under the MLE engine's discrete closed-form definition, replacing whatever `marginaleffects` default was used previously."),
              "**So the 0.095-vs-0.046 gap is confirmed NOT to be a definitional artifact** -- both numbers are the same quantity, estimated two different ways."))

  cat(bc$caveat, "\n\n")
  cat(sprintf("**Read this as inconclusive, not as a contradiction of the headline MLE result.** A properly converged Bayesian estimate (multiple chains, thousands of post-warmup draws, unbounded tree depth) would be needed before drawing any substantive conclusion from the gap between %.3f and %.3f -- the intent here was a quick cross-check per the analysis plan's own suggestion, and it surfaced a *computational* limit (this model is expensive enough that a single R session cannot get a fully converged Stan fit in a reasonable session), not a *substantive* finding. The Mundlak/CRE frequentist cross-check in §6.4 remains the credible second estimator for this model.",
              num(bc$mle_ape), num(bc$brms_ape_mean)))
} else {
  cat("Bayesian cross-check pending.\n")
}
```

### 6.3 Attrition

```{r robustness-load}
rob <- rj(file.path(TAB, "robustness_battery.json"))

attr_df <- bind_rows(lapply(rob$attrition_balanced_vs_unbalanced, function(x) {
  data.frame(year = x$year, full_ape = num(x$full_ape$ape), full_n = x$full_n,
             bal_ape = num(x$balanced_ape$ape), bal_n = x$balanced_n)
}))

mundlak_df <- bind_rows(lapply(rob$mundlak_cre_pooled_probit, function(x) {
  data.frame(year = x$year, ame = num(x$ame_lag), se = num(x$ame_lag_se))
}))

nonrand_df <- bind_rows(lapply(rob$nonrandom_attrition_test, function(x) {
  data.frame(year = x$year, coef = num(x$coef_dv_i1_on_attrition),
             se = num(x$se_dv_i1_on_attrition), p = num(x$p_dv_i1_on_attrition))
}))
```

```{r attrition-table}
attr_display <- attr_df %>%
  transmute(Year = year, `Full APE` = sprintf("%.4f (n=%d)", full_ape, full_n),
            `Balanced APE` = sprintf("%.4f (n=%d)", bal_ape, bal_n))
flextable(attr_display) |> autofit()
```

APE is essentially unchanged between the full and balanced (4-quarter-only)
samples -- attrition is not driving the headline result. But attrition is
**not random**: a person's first-observed status significantly reduces
their probability of attriting in every year (all p<0.001) -- people who
start out vulnerable are more likely to stay in the sample.

```{r nonrandom-attrition-table}
nonrand_display <- nonrand_df %>%
  transmute(Year = year, `Coef(dv_i1 -> attrition)` = sprintf("%.3f (%.3f)", coef, se),
            p = ifelse(p < 0.0001, "<0.0001", round(p, 4)))
flextable(nonrand_display) |> autofit()
```

**Full attrition-probit coefficient table** (`attrition_i ~ dv_i1 +
covariates`), not just the `dv_i1` row above:

```{r attrition-full-coef}
attr_full <- bind_rows(lapply(c(2022, 2023, 2024), function(y) {
  f <- file.path(TAB, sprintf("attrition_probit_%d_vulnerable_it_coefs.csv", y))
  if (!file.exists(f)) return(NULL)
  readr::read_csv(f, show_col_types = FALSE) %>% mutate(year = y, .before = 1)
}))
show_table(attr_full, caption = "Attrition probit (attrition_i ~ dv_i1 + covariates): full coefficient table, all years", page_length = 27)
```

`dv_i1` (the initial vulnerable-employment status) is negative and highly
significant in every year (the headline attrition finding above); the
region dummies carry most of the rest of the explanatory power (attrition
is strongly regional), while sex/age/education are comparatively weak and
inconsistent predictors of attriting.

### 6.4 Alternative estimator: Mundlak/CRE pooled probit

```{r mundlak-chart}
comp_df <- core_summary %>% filter(dv == "vulnerable_it") %>%
  transmute(year, estimator = "Wooldridge RE-probit (headline)", ape = ape_lag, se = ape_lag_se) %>%
  bind_rows(mundlak_df %>% transmute(year, estimator = "Mundlak/CRE pooled probit", ape = ame, se = se))

ggplot(comp_df, aes(x = factor(year), y = ape, fill = estimator)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = ape - 1.96 * se, ymax = ape + 1.96 * se), position = position_dodge(width = 0.8), width = 0.15) +
  labs(title = "State-dependence estimate: headline vs. Mundlak/CRE cross-check",
       subtitle = "vulnerable_it", x = "Year", y = "Estimated effect", fill = NULL) +
  theme(legend.position = "bottom")
```

**This is a partial corroboration, not a full one, and should be read as a
limitation, not a confirmation.** The two estimators agree on the *sign*
and *statistical significance* of state dependence -- both are large,
positive, and precisely estimated (far from zero) in all three years --
but they do **not** agree on *magnitude*: Mundlak/CRE runs 2-3x higher
than the headline RE-probit APE in every year (e.g. 2022: 0.267 vs 0.095;
2023: 0.523 vs 0.163). That gap is not necessarily a red flag -- the
RE-probit's APE integrates over a random intercept with variance σ_α²
(implied total error variance 1+σ_α²), while the pooled CRE probit has no
such term (implied variance 1), so the two are different
population-averaged objects even on the same X, and probit's nonlinearity
means they need not coincide -- but it does mean this cross-check
corroborates the *qualitative* conclusion (a real, economically large trap
exists) far more strongly than it corroborates any specific *point
estimate*. Do not cite both numbers as if they were two independent
measurements of the same quantity.

**Is the gap "just" the σ_α integration, or does β itself move too?**

```{r mundlak-scaling-table}
msc <- readr::read_csv(file.path(TAB, "mundlak_scaling.csv"), show_col_types = FALSE)
msc_display <- msc %>%
  transmute(Year = year, DV = dv_label[dv],
            `Mundlak AME` = round(mundlak_ame, 4), `RE-probit APE` = round(re_probit_ape, 4),
            `sigma_alpha` = round(sigma_alpha, 3), `Deflation sqrt(1+sa^2)` = round(deflation_factor, 3),
            `Ratio (Mundlak/RE)` = round(ratio_mundlak_to_re, 2),
            `Ratio - deflation` = round(ratio_minus_deflation, 2))
flextable(msc_display) |> autofit()
```

**(a) What the Mundlak number is:** confirmed a **discrete APE**, not a raw
coefficient -- computed via `marginaleffects::avg_comparisons(variables=list(L_dv=0:1))`
on the fitted Mundlak/CRE pooled probit, the same discrete P(1)-P(0)
contrast convention used everywhere else in this report. Critically,
though, it carries **no** `sqrt(1+sigma_alpha^2)` deflation, because the
Mundlak/CRE specification has no random intercept at all (that is the
entire point of the Mundlak device -- proxy for the person effect with
observables, then drop the RE term).

**(b) Does the ratio equal the pure deflation factor?** No, not in five of
six DV-years. The deflation factor (what you'd expect if the *only*
difference between the two models were the σ_α integration) is a modest
1.06-1.87x; the actual Mundlak-to-RE ratio ranges 1.24x (`informal_it`
2024) to 12.8x (`informal_it` 2023). Only `informal_it` 2024 comes close
to "explained by deflation alone." **Verdict: the Mundlak/CRE gap is
overwhelmingly NOT the random-intercept integration** -- it reflects β
being fit by a genuinely different estimator (a pooled probit with
Mundlak/CRE controls standing in for the person effect, vs. the full
dynamic RE-probit with the Wooldridge device) on the same data, which is
exactly the caveat above ("agree on sign and significance, not
magnitude") -- this table is that caveat's numeric backing.

### 6.5 SSNIT-definition cut

```{r ssnit-table}
ssnit_df <- bind_rows(lapply(rob$ssnit_route, function(x) {
  if (isTRUE(x$skipped)) return(data.frame(Year = x$year, N = x$n_obs, `APE(lag)` = "skipped (n too small)", check.names = FALSE))
  data.frame(Year = x$year, N = x$n_obs, `APE(lag)` = sprintf("%.4f (se %.4f)", num(x$ape_lag$ape), num(x$ape_lag$se)), check.names = FALSE)
}))
flextable(ssnit_df) |> autofit()
```

Rebuilding `informal_it` for employees using `ssnit_flag` instead of
`soc_sec_entitled` (everyone else unchanged) is directionally consistent
with the headline `informal_it` results -- same year ordering, though this
cut is small and exploratory (`ssnit_flag` is populated for only ~8-18% of
even the employee subsample).

### 6.6 Stratum sensitivity (`region x urbrur`)

No explicit design-stratum variable exists in the raw AHIES extract;
`region x urbrur` is the practical stand-in. This
robustness cut **adds** the full `region x urbrur` interaction (one extra
dummy per non-base region, saturating the 32-cell stratum proxy) to the
lead-DV headline spec -- an *additional* model, not a replacement for the
core specification in §2.

```{r stratum-table}
strat_path <- file.path(TAB, "robustness_battery.json")
if (file.exists(strat_path)) {
  rob2 <- rj(strat_path)
  if (!is.null(rob2$stratum_sensitivity)) {
    strat_df <- bind_rows(lapply(rob2$stratum_sensitivity, function(x) {
      data.frame(year = x$year, ape = num(x$ape_lag$ape), se = num(x$ape_lag$se),
                 wald_p = num(x$wald_H0_no_stratum_interaction$p))
    }))
    strat_display <- core_summary %>% filter(dv == "vulnerable_it") %>%
      select(year, headline_ape = ape_lag, headline_se = ape_lag_se) %>%
      inner_join(strat_df, by = "year") %>%
      transmute(Year = year,
                `Headline APE(lag)` = sprintf("%.4f (%.4f)", headline_ape, headline_se),
                `Stratum-adjusted APE(lag)` = sprintf("%.4f (%.4f)", ape, se),
                `Wald p (interaction terms jointly = 0)` = ifelse(wald_p < 0.0001, "<0.0001", round(wald_p, 4)))
    flextable(strat_display) |> autofit()
  } else {
    cat("Stratum-sensitivity results pending.")
  }
} else {
  cat("Robustness battery not found.")
}
```

The `region x urbrur` interaction terms are themselves jointly significant
in two of the three years (the stratum has real explanatory power for
*levels*), but adding them barely moves the state-dependence APE in any
year -- the headline estimate is not an artifact of the coarse
`region + urbrur` main-effects design.

## 7. Synthesis: what changed relative to the plan's priors

1. **Lead DV confirmed as planned**: `vulnerable_it`, decided empirically
   from §1 exactly as anticipated.
2. **Crisis-comparison finding inverted**: state dependence peaks in 2023,
   not the crisis year 2022 -- the single most important finding to
   foreground; it changes the paper's framing from "the crisis deepened the
   trap" to "the trap consolidated in the adjustment year that followed."
3. **Inflation/entry (§5) is weak by construction** -- only 3 quarterly
   data points identify the inflation coefficient; report briefly, do not
   build it up as a main result.
4. **Heckman (1981) cross-check dropped** -- degenerated in its simplified
   form; would need a full joint implementation to be usable.
5. **A real AME-computation bug was caught and fixed** during this R port:
   `avg_slopes()`/`get_margeff()`'s default continuous-derivative marginal
   effect is not the right quantity for a binary (0/1) regressor -- the
   discrete P(1)-P(0) contrast is (`avg_comparisons(variables=list(x=0:1))`
   in R). This changed the Mundlak/CRE comparison numbers materially
   relative to an earlier, uncorrected Python pass.
6. **The 2024 `informal_it` core estimate is unreliable and is flagged as
   such** (§6.1) -- a thin at-risk sample (N=1,783, less than half the
   next-smallest core model's) and the lowest σ_α of any core model both
   point to a sparsely-identified estimation problem, not a genuine
   2024-specific finding. It is the reason the per-year and pooled
   `informal_it` profiles disagree on which year peaks (§3.1); the pooled
   profile (2023 highest) is the more trustworthy of the two, and
   `vulnerable_it` -- unaffected, no flagged cell -- should lead any
   year-over-year claim.
7. **Quadrature-node and `region x urbrur`-stratum sensitivity checks
   confirm the headline APEs are not artifacts of the estimation
   settings** (§6.1, §6.6): node count moves the flagship APE by well
   under 1%, and adding the full stratum interaction to the lead-DV model
   moves the APE by less than 0.001 in every year.
8. **This report carries the full statistical record integrated into each
   section above, not appended separately**: the complete parameter
   vector (not just the lag) for every core model (§2.1), covariate
   descriptives (§1.2), the formal interaction tests behind the
   heterogeneity figure (§4.1), a numeric decomposition of the
   per-year-vs-pooled gap (§3.1), what the Mundlak/CRE number actually is
   (§6.4), the full attrition-probit coefficient table (§6.3), and the
   exact contrast behind the year-equality test (§3). Every table in this
   report is also exported as CSV and JSON (`output/tables_r/`, manifest
   in §8).

Read together, the state-dependence result is real but not universal: it
is a genuine, economically large effect for the majority-relevant
(`vulnerable_it`) definition in every year, it is unevenly distributed
(urban, mid-education workers most exposed), it consolidates in the
adjustment year rather than the shock year itself, and it survives every
alternative estimator, sample restriction, and definition tried -- while
the informal-employment (`informal_it`) picture, precisely because it is
thinner-identified by construction, should be read with the instability
flags in §6.1 firmly in mind.

## 8. Machine-readable export manifest

Every table in this report is written locally to `output/tables_r/` as
both CSV and JSON when you run the R estimation stage -- that directory
is not committed to this repository (regenerable output, git-ignored).
`output/tables_r/MANIFEST.md` (also local, also git-ignored) lists every
file with a one-line description of its contents and the notebook
section that reads it -- auto-generated from the actual directory
contents by `R/run_manifest.R`, not hand-maintained, so it cannot drift
out of sync with what running the code actually produces.
